Estimating Planetary Habitability via Particle Swarm Optimization of CES Production Functions.

### 2018-01-14 docs/report/draft.tex
Finished up to CO by PSO.

\documentclass{article}

\usepackage[utf8]{inputenc}
\usepackage{mathtools}

% Information for header.
\title{Optimizing the Cobb Doughlas Habitability Production function using
Particle Swarm Optimization}
\author{Abhijit J.\ Theophilus}
\date{\today}

@@

@@

\section{Introduction}
% Note to self, improve the flow of the report.
% Find citations.
Under most paradigms, optimization of a continuous function requires the calculation of a gradient. This might not
always be feasible for non-polynomial functions in high-dimensional search spaces. This is further complicated when the
optimization must be performed under constraints. \pso\ is a metaheuristic for continuous function optimization that does
away with the need for a gradient. It employs a large number of particles that traverse the search space converging
toward a global best solution encountered by at least one of the particles.

\pso\ is a distributed method that requires simple mathematical operators and few lines of code, making it a lucrative
solution where computation resources are at a premium. Its implementation is highly parallelizable. It scales with the
dimensionality of the search space. Constrained optimization can be incorporated within the standard algorithm through
the use of a penalty function.

% Pull out citations from the Applications paper.
\pso\ has been successfully adapted to a wide range of design optimization problems, e.g., networks and VLSI design.
\pso\ has also found applications in machine learning under clustering, feature detection and classification.  As a
modelling paradigm, it has been used for constructing customer satisfaction models, MIDI music models, friction models,
etc.

In this paper, \pso\ has been applied to estimate the Cobb Douglas Habitability Score of an exoplanet. The score is
estimated by maximizing the Cobb-Douglas Habitability Production Function. The function considers four parameters,
radius, mass, escape velocity and surface temperature, taken from the Exoplanets Catalog hosted by the Planetary
Habitability Laboratory, UPR Arecibo. Since the Cobb-Douglas function is concave under certain constraints, the
problem can be reduced to a convex optimization over the negative of the production function.

\section{Cobb-Douglas Habitability Score}

\subsection{Cobb-Douglas Habitability Production Function}
The general form of the Cobb-Douglas production function is given by,
\begin{equation}
  Y = k\cdot{(x_1)}^\alpha\cdot{(x_2)}^\beta,
\end{equation}
where $k$ is a constant that may be set according to requirement, $x_1$ and $x_2$ are the input parameters to the
function, $Y$ is the total output production value, and $\alpha$ and $\beta$ are called the elasticity coefficients.

% \begin{pointers}
% \item Describe the Cobb Douglas Habitability Score.
% \item Describe CRS, DRS and IRS\@.
% \item Splitting the CDHS score into two values.
% \item State the equations.
% \item Score is estimated by maximizing the CDHPF fn.
% \end{pointers}

\section{Particle Swarm Optimization}
\begin{pointers}
\item Equations. Modification for max.
\item Constrained optimization modifications.
\item Algorithm for PSO\@.
\end{pointers}

>>>>

\documentclass[10pt]{article}

\usepackage[margin=1.2in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}

\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage{mathtools}

% Information for header.
\title{Estimating Planetary Habitability via Particle Swarm Optimization of CES Production Functions.}
\author{Abhijit J.\ Theophilus}
\date{\today}

@@

% For algorithms
\floatname{algorithm}{Procedure}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

% For math
\DeclareMathOperator*{\argmin}{arg\,min}

@@

\section{Introduction}\label{sec:intro}
% Note to self, improve the flow of the report.
% Find citations.
The search for extra-terrestial life and potentially habitable extrasolar planets has been an international venture
demanding large investments in cost and effort, since Frank Drake's attempt with Project Ozma in the mid-20th century.
The first exoplanet was officially confirmed in 1992 which marked the start of a trend that has lasted 25 years and
yielded over 3,700 confirmed exoplanets. There have been attempts to assess the habitability of these planets and to
assign a score based on their similarity to Earth. Two such habitability scores are the Cobb-Douglas Habitability Score
(CDHS) and the Constant Elasticity Earth Similarity Approach (CEESA) score. Estimating these scores involves maximizing
a production function subjected to a set of constraints on its input variables.

Under most paradigms, maximizing a continuous function requires calculating a gradient. This may not always be feasible
for non-polynomial functions in high-dimensional search spaces. Further, subjecting the input variables to constraints,
as needed by CDHS and CEESA, are not always be straightforward to represent within the model. This paper presents an
approach to Constrained Optimzation (CO) using the swarm intelligence metaheuristic. \pso\ (PSO) is a method for
optimizing a continuous function that does away with the need for a gradient. It employs a large number of particles
that traverse the search space converging toward a global best solution encountered by at least one of the particles.

\pso\ is a distributed method that requires simple mathematical operators and short segments of code, making it a
lucrative solution where computational resources are at a premium. Its implementation is highly parallelizable. It
scales with the dimensionality of the search space. The standard PSO algorithm does not deal with constraints, but
through variations in initializing and updating particles, constraints are straightforward to represent and adhere to,
as seen in Section~\ref{subsec:copso}.

% Pull out citations from the Applications paper.
PSO has been adapted to a wide range of design optimization problems, e.g., networks and VLSI design. It has found
applications in machine learning under clustering, feature detection and classification. As a modeling paradigm, it has
been used for constructing customer satisfaction models, modeling MIDI music and chaotic time series modeling.

This paper demonstrates the applicability of \pso\ in estimating CDHS and CEESA scores of an exoplanet by maximizing
their respective production functions, discussed in Sections~\ref{subsec:cdhs} and~\ref{subsec:ceesa}. CDHS considers
the planet's Radius, Mass, Escape Velocity and Surface Temperature, while CEESA includes a fifth parameter, the Orbital
Eccentricity of the planet. The Exoplanets Catalog hosted by the Planetary Habitability Laboratory, UPR Arecibo records
these parameters for each exoplanet in Earth Units. Section~\ref{sec:results} reports the performance of PSO and
contrasts it against an earlier effort to estimate these scores using Stochastic Gradient Ascent (SGA).

\section{Habitability Scores}
% Add the citations to Saha's papers.

\subsection{Cobb-Douglas Habitability Score}\label{subsec:cdhs}
Estimating the Cobb-Douglas Habitability Score (CDHS) requires estimating an interior CDHS (CDHS\textsubscript{i}) and a
surface CDHS (CDHS\textsubscript{s}) by maximizing the following production functions,
\begin{subequations}
  \begin{alignat}{4}
    Y_i\ &=\ {CDHS}_i\ &=&\ R^\alpha.&D^\beta\label{eq:cdhsi}\\
    Y_s\ &=\ {CDHS}_s\ &=&\ {V_e}^\gamma.&{T_s}^\delta\label{eq:cdhss}
  \end{alignat}
\end{subequations}
where, $R$, $D$, $V_e$ and $T_s$ are density, radius, escape velocity and surface temperature respectively. $\alpha$,
$\beta$, $\gamma$ and $\delta$ are the elasticity coefficients subject to,
\begin{equation}
  0 < \alpha,\beta,\gamma,\delta < 1
\end{equation}
Equations~\ref{eq:cdhsi} and~\ref{eq:cdhss} are convex when the returns to scale are either constant or decreasing,
marked by two constraints on the elasticity coefficients,
\begin{subequations}
  \begin{align}
    \alpha+\beta = 1,\quad\gamma+\delta = 1 && \text{for Constant Returns to Scale (CRS),}\\
    \alpha+\beta < 1,\quad\gamma+\delta < 1 && \text{for Decreasing Returns to Scale (DRS).}
  \end{align}
\end{subequations}
The final CDHS is the convex combination of the interior and surface CDHS values as given by,
\begin{equation}
  Y\ =\ w_i.Y_i + w_s.Y_s
\end{equation}

\subsection{Constant Elasticity Earth Similarity Approach}\label{subsec:ceesa}
The Constant Elasticity Earth Similarity Approach (CEESA) uses the following production function to estimate the
habitability score of an exoplanet,
\begin{equation}\label{eq:ceesa}
  Y = {(r.R^\rho+d.D^\rho+t.{T_s}^\rho+v.{V_e}^\rho+e.E^\rho)}^{\eta/\rho}
\end{equation}
where, $E$ is the fifth parameter denoting Orbital Eccentricity. The value of $\rho$ lies within $0<\rho\leq 1$.
The coefficients ($r$, $d$, $t$, $v$ and $e$) are constrained by,
\begin{subequations}
  \begin{align}
      0 < r,d,t,v,e < 1\\
      r+d+t+v+e = 1
  \end{align}
\end{subequations}
The value of $\eta$ is constrained by the scale of production used,
\begin{subequations}
  \begin{align}
    0 < \eta < 1 && \text{for Constant Returns to Scale (CRS),}\\
    \eta = 1 && \text{for Decreasing Returns to Scale (DRS).}
  \end{align}
\end{subequations}

\section{Particle Swarm Optimization}\label{sec:pso}
Particle Swarm Optimization (PSO) is a biologically inspired metaheuristic for finding the global minima of a function.
Traditionally designed for unconstrained inputs, it works by iteratively converging a population of randomly initialized
solutions, called particles, toward a globally optimal solution. Each particle in the population keeps track of its
current position and the best solution it has encountered so far, called $pbest$. Each particle also has an associated
randomized velocity used to traverse the search space. The swarm keeps track of the overall best solution, called
$gbest$. Each iteration of the swarm updates the velocity of the particle towards its $pbest$ and the $gbest$ values.

\subsection{PSO for Unconstrained Optimation}\label{subsec:uopso}
Let $f(x)$ be the function to be minimized, where $x$ is a $d$-dimensional vector. $f(x)$ is also called the fitness
function. Algorithm~\ref{alg:unop} outlines the approach to minimizing $f(x)$ using PSO.\@ A set of particles are randomly
initialized with a position and a velocity, where $l$ and $u$ are the lower and upper boundaries of the search space.
The position of the particle corresponds to its associated solution. The algorithm initializes each particle's $pbest$
to its initial position. The $pbest$ position that corresponds to the minimum fitness is selected to be the $gbest$
position of the swarm.

On each iteration, the algorithm updates the velocity and position of each particle. For each particle, it picks two
random numbers $u_g, u_p$ from a uniform distribution, $U(0,1)$ and updates the particle velocity as indicated in
line~\ref{algline:vup}. Here, $\mu$ is the friction coefficient and $\lambda_g,\lambda_p$ are the global and particle
learning rates. If the new position of the particle corresponds to a better fit than its $pbest$, the algorithm updates
$pbest$ to the new position. Once the algorithm has updated all particles, it updates $gbest$ to the new overall best
position. A suitable termination criteria for the swarm, under convex optimization, is  when the $gbest$ position has
not changed by the end of the iteration.

\begin{algorithm}
  \begin{algorithmic}[1]
    \Require{$f(x)$, the function to minimize.}
    \Ensure{global minimum of $f(x)$.}
    \For{each particle $i\gets 1,n$}
      \State{$p_i \sim U(l,u)$}
      \State{$v_i \sim U(-|u-l|, |u-l|)$}
      \State{${pbest}_i \gets p_i$}
    \EndFor\
    \State{$gbest \gets \smashoperator{\argmin\limits_{{pbest}_i, i \in {1,n}}} f({pbest}_i)$}
    \Repeat\
      \State{$oldbest \gets gbest$}
      \For{each particle $i \gets 1,n$}
        \State{$u_p, u_g \sim U(0,1)$}
        \State{$v_i \gets \mu.v_i + \lambda_g.u_g.({gbest-p_i}) + \lambda_p.u_p.({{pbest}_i-p_i})$}\label{algline:vup}
        \State{$p_i \gets p_i + v_i$}
        \If{$f(p_i) < f({pbest}_i)$}
          \State{${pbest}_i \gets p_i$}
        \EndIf\
      \EndFor\
      \State{$gbest \gets \smashoperator{\argmin\limits_{{pbest}_i, i \in {1,n}}} f({pbest}_i)$}
    \Until{$|oldbest - gbest| < threshold$}
    \State{\textbf{return} {$f(gbest)$}}
  \end{algorithmic}
  \caption{Algorithm for PSO.}\label{alg:unop}
\end{algorithm}

\subsection{PSO with Leaders for Constrained Optimization}\label{subsec:copso}
Although PSO has eliminated the need to estimate the gradient of a function, as seen in Section~\ref{subsec:uopso}, it
still is not suitable for constrained optimization. The standard PSO algorithm does not ensure that the initial
solutions are feasible, and neither does it guarantee that the individual solutions will converge to a feasible global
solution. Solving the initialization problem is straightforward, resample each random solution from the uniform
distribution until every initial solution is feasible. To solve the convergence problem, each particle uses another
particle's $pbest$ value, called $lbest$, instead of its own to update its velocity. Algorithm~\ref{alg:cop} describes
this process.

On each iteration, for each particle, the algorithm first picks two random numbers $u_g,u_p$ as before. It then selects
a $pbest$ value from all particles in the swarm that is closest to the position of the particle being updated as its
$lbest$. The $lbest$ value substitutes ${pbest}_i$ in the velocity update equation. While updating $pbest$ for the
particle, the algorithm checks if the current fit is better than $pbest$, and performs the update if the current
position satisfies all constraints.

\begin{algorithm}
  \begin{algorithmic}[1]
    \Require{$f(x)$, the function to minimize.}
    \Ensure{global minimum of $f(x)$.}
    \For{each particle $i\gets 1,n$}
      \Repeat
        \State{$p_i \sim U(l,u)$}
      \Until{$p_i$ satisfies all constraints}
      \State{$v_i \sim U(-|u-l|, |u-l|)$}
      \State{${pbest}_i \gets p_i$}
    \EndFor\
    \State{$gbest \gets \smashoperator{\argmin\limits_{{pbest}_i, i \in {1,n}}} f({pbest}_i)$}
    \Repeat\
      \State{$oldbest \gets gbest$}
      \For{each particle $i \gets 1,n$}
        \State{$u_p, u_g \sim U(0,1)$}
        \State{$lbest \gets \smashoperator{\argmin\limits_{{pbest}_j, j \in {1,n}}} {\|{pbest}_j -
            p_i\|}^2$}\label{algline:lbest}
        \State{$v_i \gets \mu.v_i + \lambda_g.u_g.({gbest-p_i}) + \lambda_p.u_p.({lbest-p_i})$}\label{algline:cvup}
        \State{$p_i \gets p_i + v_i$}
        \If{$f(p_i) < f({pbest}_i)$ \textbf{and} $p_i$ satisfies all constraints}
          \State{${pbest}_i \gets p_i$}
        \EndIf\
      \EndFor\
      \State{$gbest \gets \smashoperator{\argmin\limits_{{pbest}_i, i \in {1,n}}} f({pbest}_i)$}
    \Until{$|oldbest - gbest| < threshold$}
    \State{\textbf{return} {$f(gbest)$}}
  \end{algorithmic}
  \caption{Algorithm for CO by PSO.}\label{alg:cop}
\end{algorithm}
